Modal Abstractions for Virtualizing Memory Addresses

### 2025-07-21 background.tex
+

On the x86-64 architecture, the \textsc{MMU}'s address translation uses a sparse hierarchical set of tables:
\emph{page tables} (referring to pages of memory). As Figure \ref{fig:pagetables} (based on Figure 5-17 of the 
AMD64 architecture manual~\cite{amd64_manual_vol2}\footnote{While x86 up through its 32-bit incarnation were due to Intel,
the x86-64 architecture as a 64-bit extension to x86 was originally due to AMD. As a result, it is sometimes also referred to as the \texttt{amd64} architecture.}) shows a \emph{page-table walk} for a virtual address. The page tables are managed by the OS, typically by a \emph{virtual memory manager} (VMM).\footnote{Not to be confused with Virtual 
Machine Monitor. We focus on non-hypervisor scenarios, but hardware virtualization extensions for both 
x86-64 and ARM make use of an additional set of page tables translating what a \emph{guest} considers to be its 
(virtualized) physical memory to actual physical memory. Our contributions should offer value in this scenario as well.}
Typically each process has its own 
page table, which the OS registers with the CPU by storing the (page-aligned)
physical address of the root of the page table tree (the start of the L4 table shown as \mathsf{root} in Figure \ref{fig:pagetables}) in a specific register (\texttt{cr3}) as part of switching to a new process. 
Using different mappings, which map only disjoint portions of physical memory (with some exceptions in the next section) 
is how the OS ensures memory isolation between processes.

In Figure \ref{fig:pagetablescode}, we see a simplified kernel code, \lstinline|walkpgdir|, implementing this page-table walk for translating the virtual address $\vaddr$ over the page-table tree with unique root address $\mathsf{root}$ shown in Figure \ref{fig:pagetables}.
As shown in Figure at the top of Figure \ref{fig:pagetables}, on x86-64, standard configurations use 4 levels of page tables, labelled levels 4 through 1, with lookups in the level 
1 page table resulting in the actual page of physical memory holding the requested data, and the low-order 12 bits 
being used to index into this page.\footnote{Technically levels 1--3 have explicit historical names,
but for brevity and consistency, we simply number them, in keeping with the newer 5th level. Our formalization only deals with 4-level page tables, but is straightforwardly extensible to 5.} 
As depicted at the top of Figure \ref{fig:pagetables}, for each level of translation, address-translation implementation repeatedly takes designated slices of the virtual address and indexing into a table --- e.g., L4Offset(va) at Line 2 of \lstlinline|walkpgdir| in Figure \ref{fig:pagetablescode}. 
The final lookup in the page tables gives the base physical address of a 4KB page of physical memory, to which the low-order
bits of the accessed virtual address are added to determine the actual physical address retrieved. The entries (e.g., L4 Entry in Figure \ref{fig:pagetables}) of each table are 64 bits wide, but each points to a physical address aligned to 4KB (4096 byte) boundaries, which leaves 12 bits to spare to 
control a validity bit (called the \emph{present} bit), a read-write bit (which permits write access through the entry if and only if it is set), and a range of additional bits which can be used to control caching, write-through, and more. This paper will only consider the present bit (0). As we see in Figure \ref{fig:pagetables}, the entry L2 is not be present (shown as dashed in Figure \ref{fig:pagetables}). In order to able to access to the subsequent table base address (e.g., $\mathsf{l1}$ in Line 7 of \lstinline|walkpgdir| in Figure \ref{fig:pagetablescode}). Accessing to the next table is realized in \lstinline|pte_get_next_table| in Figure \ref{fig:pagetablescode} to which the current entry (e.g., $\mathsf{l2_entry}$ in Line 7 of \lstinline|walkpgdir| in Figure \ref{fig:pagetablescode}), and this method first check if the entry is present (Line 7 of \lstinline|pte_get_next_table| in Figure \ref{fig:pagetablecodes}). For a nonexisting entry such as L2 entry, the execution goes through allocation path shown in Lines 7-15 of \lstinline|pte_get_next_table| in Figure \ref{fig:pagetablecodes} which allocates a \emph{physical addresses} (512 L1 entries) and make $\mathsf{l2}$ pointer's frame address to show the first entry of the L1 entries fresly allocated (Line 12 of \lstinline|pte_get_next_table| in Figure \ref{fig:pagetablecodes}), and set the present bit of $\mathsf{l2_entry}$ (Line 12 of \lstinline|pte_get_next_table| in Figure \ref{fig:pagetablecodes}). After the validity of page-table walk is restored, \lstinline|pte_get_next_table| converts the physical frame address of the entry to a \emph{virtual address} in Lines 16-20 of \lstinline|pte_get_next_table| in Figure \ref{fig:pagetablecodes} return the virtual address of the subsequent ($\mathsf{next}$ in in Figure \ref{fig:pagetablecodes}) base address (e.g., $\mathsf{l1}$ in Line 7 of \lstinline|walkpgdir| in Figure \ref{fig:pagetablescode}).

>>>>

On the x86-64 architecture, the \textsc{MMU}'s address translation uses a sparse hierarchical set of tables:
\emph{page tables} (referring to pages of memory). As Figure \ref{fig:pagetables} (based on Figure 5-17 of the 
AMD64 architecture manual~\cite{amd64_manual_vol2}\footnote{While x86 up through its 32-bit incarnation were due to Intel,
the x86-64 architecture as a 64-bit extension to x86 was originally due to AMD. As a result, it is sometimes also referred to as the \texttt{amd64} architecture.}) shows a \emph{page-table walk} for a virtual address. The page tables are managed by the OS, typically by a \emph{virtual memory manager} (VMM).\footnote{Not to be confused with Virtual 
Machine Monitor. We focus on non-hypervisor scenarios, but hardware virtualization extensions for both 
x86-64 and ARM make use of an additional set of page tables translating what a \emph{guest} considers to be its 
(virtualized) physical memory to actual physical memory. Our contributions should offer value in this scenario as well.}
Typically each process has its own 
page table, which the OS registers with the CPU by storing the (page-aligned)
physical address of the root of the page table tree (the start of the L4 table shown as $\mathsf{root}$ in Figure \ref{fig:pagetables}) in a specific register (\texttt{cr3}) as part of switching to a new process. 
Using different mappings, which map only disjoint portions of physical memory (with some exceptions in the next section) 
is how the OS ensures memory isolation between processes.

In Figure \ref{fig:pagetablescode}, we see a simplified kernel code, \lstinline|walkpgdir|, implementing this page-table walk for translating the virtual address $\vaddr$ over the page-table tree with unique root address $\mathsf{root}$ shown in Figure \ref{fig:pagetables}.
As shown in Figure at the top of Figure \ref{fig:pagetables}, on x86-64, standard configurations use 4 levels of page tables, labelled levels 4 through 1, with lookups in the level 
1 page table resulting in the actual page of physical memory holding the requested data, and the low-order 12 bits 
being used to index into this page.\footnote{Technically levels 1--3 have explicit historical names,
but for brevity and consistency, we simply number them, in keeping with the newer 5th level. Our formalization only deals with 4-level page tables, but is straightforwardly extensible to 5.} 
As depicted at the top of Figure \ref{fig:pagetables}, for each level of translation, address-translation implementation repeatedly takes designated slices of the virtual address and indexing into a table --- e.g., L4Offset(va) at Line 2 of \lstinline|walkpgdir| in Figure \ref{fig:pagetablescode}. 
The final lookup in the page tables gives the base physical address of a 4KB page of physical memory, to which the low-order
bits of the accessed virtual address are added to determine the actual physical address retrieved. The entries (e.g., L4 Entry in Figure \ref{fig:pagetables}) of each table are 64 bits wide, but each points to a physical address aligned to 4KB (4096 byte) boundaries, which leaves 12 bits to spare to 
control a validity bit (called the \emph{present} bit), a read-write bit (which permits write access through the entry if and only if it is set), and a range of additional bits which can be used to control caching, write-through, and more. This paper will only consider the present bit (0). As we see in Figure \ref{fig:pagetables}, the entry L2 is not be present (shown as dashed in Figure \ref{fig:pagetables}). In order to able to access to the subsequent table base address (e.g., $\mathsf{l1}$ in Line 7 of \lstinline|walkpgdir| in Figure \ref{fig:pagetablescode}). Accessing to the next table is realized in \lstinline|pte_get_next_table| in Figure \ref{fig:pagetablescode} to which the current entry (e.g., $\mathsf{l2_entry}$ in Line 7 of \lstinline|walkpgdir| in Figure \ref{fig:pagetablescode}), and this method first check if the entry is present (Line 7 of \lstinline|pte_get_next_table| in Figure \ref{fig:pagetablecodes}). For a nonexisting entry such as L2 entry, the execution goes through allocation path shown in Lines 7-15 of \lstinline|pte_get_next_table| in Figure \ref{fig:pagetablecodes} which allocates a \emph{physical addresses} (512 L1 entries) and make $\mathsf{l2}$ pointer's frame address to show the first entry of the L1 entries fresly allocated (Line 12 of \lstinline|pte_get_next_table| in Figure \ref{fig:pagetablecodes}), and set the present bit of $\mathsf{l2_entry}$ (Line 12 of \lstinline|pte_get_next_table| in Figure \ref{fig:pagetablecodes}). After the validity of page-table walk is restored, \lstinline|pte_get_next_table| converts the physical frame address of the entry to a \emph{virtual address} in Lines 16-20 of \lstinline|pte_get_next_table| in Figure \ref{fig:pagetablecodes} return the virtual address of the subsequent ($\mathsf{next}$ in in Figure \ref{fig:pagetablecodes}) base address (e.g., $\mathsf{l1}$ in Line 7 of \lstinline|walkpgdir| in Figure \ref{fig:pagetablescode}).
